---
title: "field-vs-lab-doses-r"
author: "Jake Martin"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_document:
    code_download: true
    code_folding: hide
    depth: 4
    number_sections: no
    theme:  cosmo
    toc: yes
    toc_float: yes
    toc_depth: 4
  pdf_document:
    toc: yes
knit: |
  (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = paste0(
       'index.html'
      ),
      envir = globalenv()
    )
  })
---

#########################
# README
#########################

**ARTICLE** <br>
This R script is associated with **Martin et al (2025)** "Aligning Behavioural Ecotoxicology Tests with Real-World Water Concentrations: Current Minimum Tested Levels Far Exceed Environmental Reality" DOI:
<br>

**AUTHORS**<br>
Jake M. Martin <sup>1,2,3</sup>* <br>
Jack A. Band <sup>2,4</sup> <br>
Erin S. McCallum <sup>1</sup> 
<br>

**AFFILIATIONS** <br>
(1) School of Life and Environmental Sciences, Deakin University, Geelong, Australia <br>
(2) Department of Wildlife, Fish, and Environmental Studies, Swedish University of Agricultural Sciences, Umeå, Sweden <br>
(3) School of Biological Sciences, Monash University, Melbourne, Australia <br>
(4) Institute of Zoology, Zoological Society of London, London, United Kingdom <br>
(*) Corresponding author 
<br>

**AIM** <br>
The aim of this study is two-fold: (1) to compare the target pharmaceuticals tested in behavioural ecotoxicology to what have been detected in the environment, highlighting potential targets for future investigation; (2) identify whether tested concentrations in behavioural ecotoxicology studies reflect those that are detected in surface waters and/or waste water.
<br>


**METHODS** <br>
To achieve these aims, this study capitalises on four extensive open access databases, one of which focuses on peer-review research on the behavioural ecotoxicology of pharmaceuticals<sup>1</sup>, with the remaining three focusing on pharmaceutical environmental occurrence and concentrations<sup>2,3,4</sup>.<br>

(1) The Evidence of the Impacts of Pharmaceuticals on Aquatic Animal Behaviour (EIPAAB) database (Martin et al. 2024; DOI: https://doi.org/10.32942/X2NG9R) <br>

(2) The Umwelt Bundesamt Pharmaceuticals in the environment (PHARMS-UBA) database (access link: https://www.umweltbundesamt.de/en/database-pharmaceuticals-in-the-environment-1) <br>

(3) the Wilkinson global river survey data (Wilkinson et al 2022: DOI: https://doi.org/10.1073/pnas.2113947119), specifically "Dataset S4. Database of pharmaceutical concentrations at all the sampling locations monitored in this project" <br>

(4) The NORMAN EMPODAT Database - Chemical Occurrence Data (https://www.norman-network.com/nds/empodat/chemicalSearch.php?s=new), searched for all compounds, all countries, all years, all data sources, all organisations, all laboratories,  in Surface water (Sea water, River water, Lake water, Transitional water, Coastal water, Territorial (marine) water, Reservoirs, Others) and waste water (Urban, Industrial, Municipal, Other) Ecosystems/matrices on 19-12-2024
<br>

The four databases were filtered to focus on pharmaceuticals compounds, environmental samples from surface waters/waste water (e.g. river water, lake water, sea/ocean water, reservoirs, waste water treatment plant [WWTP] effluent), and environmental/lab samples measured in units of mass per volume of water, where necessary. <br>

**RESULTS**<br>
Through the combination of these open access databases, we were able to compare the data from over **XX** behavioural ecotoxicology exposures. In total, **XX** pharmaceutical compounds tested in the EIPAAB database (with an environmental motivation and unit measures in mass per volume of water) also had associated environmental data in at least one of the three databases (XX% of possible compounds). We detected a positive association with the number of behavioural ecotoxicology tests and total number of environmental detections per compound, but not the proportion of positive detections per compound (i.e. relative number of samples above limit of detection). In terms of concentrations, on average, the concentrations used in behavioural ecotoxicology were **XX times higher** then median surface water concentrations, and **XX times higher** then those in effluent. Testing each database separately resulted in very different estimates of relative concentrations and occurrence correlations... Lastly, we highlighted a number of pharmaceutical contaminants that have high environmental detections, but have not yet been the subject of any behavioural ecotoxicology test, and thus could be key compounds to focus future reserach.  <br>

**SCRIPT** <br>
This is an R markdown script written in R studio (2023.09.0+463 “Desert Sunflower” Release). The 'concentration-conundrum' git repository hosts this script, and if downloaded (or pulled) it will reproduce all data tidy/filter, analysis, and visualisations used in Martin et al (2025). The GitHub repository includes all raw input data from the four open access databases, as well as all output data and figures. I have tried to structure this code with Open, Reliable, and Transparent (ORT) coding practices in mind. Feel free to reach out if anything is unclear.

The R script first cleans and filters each databases to make them comparable. In terms of filtering, we target pharmaceutical compounds, surface waters/waste water effluent, and environmental/lab samples measured in units of mass per volume of water (e.g. ug/L). <br>

We then summarise occurrence and concentrations between EIPAAB the three environmental databases (UBA, Wilkson and NORMAN) separately, to highlight the consistency of different environmental databases. <br>

Lastly we combined all environmental datasets for an overall assesses of occurrence and concentrations between EIPAAB and environmental data. <br>

GitHub: https://github.com/JakeMartinResearch/concentration-conundrum

#########################
# Contact
#########################

**Jake M. Martin** <br>

**Email**: jake.martin@deakin.edu.au (or jake.martin.research@gmail.com) <br>

**Web**: jake.martin.org <br>

**GitHub**: https://github.com/JakeMartinResearch <br>

#########################
# Knit settings 
#########################

Here we define our Knit settings, to make the output more user friendly, and to cache output for faster knitting.

```{r setup}
#kniter seetting
knitr::opts_chunk$set(
message = FALSE,
warning = FALSE, # no warnings
cache = TRUE,# Cacheing to save time when kniting
tidy = TRUE
)
```

#########################
# Required packages
#########################

These are the R packages required for this script. You will need to install a package called pacman to run the p_load function. 

```{r, message=FALSE, results='hide'}
# this installs and load packages
# need to install pacman
pacman::p_load("ggplot2", 
               "ggthemes", 
               "ggfortify",
               "ggridges",
               "gghalves",
               "ggExtra",
               "plotly",
               "colorspace", 
               "ggrepel", 
               "ggdist",
               "gt",
               "webshot2", # data visualisation
               
               "data.table",
               "stringr",
               "tidyverse", 
               "janitor", 
               "readxl", 
               "broom", # data tidy
               
               "brms",
               "performance" # data analysis 
                       )
```


#########################
# Custom functions 
#########################

Here are some custom function used within this script. <br>

*tidy_string* : To tidy the names in UBA database

```{r}
tidy_string <- function(string) {
  string %>%
    stringr::str_remove_all("[-()]") %>%  # Remove unwanted characters
    stringr::str_replace_all("\\s{2,}", " ") %>%  # Replace double spaces with a single space
    stringr::str_replace_all("\\s", "_") %>%  # Replace spaces with underscores
    stringr::str_replace_all("[/]", "_") %>% 
    stringr::str_replace_all("[µ]", "u") %>% 
    tolower()
}
```

*make_na* : Takes a sting and replaces with NA based on specification in the UBA database (e.g. -9999, [-], -, or "")

```{r}
make_na <- function(string) {
    dplyr::case_when(
      string == -9999 ~ NA,
      string == "[-]" ~ NA,
      string == "-" ~ NA,
      string == "" ~ NA,
      TRUE ~ string
      )
}
```

*sentence_case* : Makes a string into sentence case

```{r}
sentence_case <- function(text) {
  text <- tolower(text) # Convert the entire text to lowercase
  text <- sub("^(\\s*\\w)", "\\U\\1", text, perl = TRUE) # Capitalise the first letter
  return(text)
}
```


#########################
# Directories 
#########################

Here we define the directories for the project. We also make the output directories if they don't already exist. <br>

## Input

These are the input directors for the databases used in this project. <br>

**input_wd**: Directory for the input data

```{r}
wd <- getwd() # getwd tells us what the current wd is, we are using this to drop it in a variable called wd
input_wd <- paste0(wd, "./input-data") # creates a variable with the name of the wd we want to use
```


## Output

These are output directories for data and figures.<br>

**output_wd**: Directory for the output data

```{r, echo = TRUE, results = "hide"}
wd <- getwd() # getwd tells us what the current wd is, we are using this to drop it in a variable called wd
output_wd <- paste0(wd, "./output-data") # creates a variable with the name of the wd we want to use
ifelse(!dir.exists("output-data"), dir.create("output-data"), "Folder already exists")
```

**output_fig_wd**: Directory for the output figures

```{r, echo = TRUE, results = "hide"}
output_fig_wd <- paste0(wd, "./output-fig")
ifelse(!dir.exists("output-fig"), dir.create("output-fig"), "Folder already exists")
```

**output_models_ws**:: Directory for the output models

```{r}
output_mod_wd <- paste0(wd, "./output-mods")
ifelse(!dir.exists("output-mods"), dir.create("output-mods"), "Folder already exists")
```

#########################
# Databases 
#########################

In this section of the script we will import the different databases and tidy them. <br>

## EIPAAB

Import the EIPAAB database, there are 901 articles and 1739 rows of data.

```{r}
setwd(input_wd)
EIPAAB <- read.csv("EIPAAB-database.csv")

articles <- EIPAAB %>% 
  dplyr::distinct(article_id) %>% 
  nrow(.)

data_points <- EIPAAB %>% 
  nrow(.)

print(paste0("n articles = ", articles, "; n data rows = ", data_points))
```

### Filtering

We are **filter** the EIPAAB database for studies with an Environmental motivation and studies with concentrations reported in mass per volume of water (ug/L). We now have 463 articles and 767 rows of data (we removed 438 articles and 972 rows of data).

```{r}
EIPAAB <- EIPAAB %>% 
  dplyr::mutate(compound_min_dose_unit_std = tidy_string(compound_min_dose_unit_std),
                compound_max_dose_unit_std = tidy_string(compound_max_dose_unit_std)) %>% 
  dplyr::filter(study_motivation == "Environmental", 
                compound_min_dose_unit_std == "ug_l", compound_max_dose_unit_std == "ug_l") %>% 
  dplyr::mutate(compound_name = tolower(compound_name),  
                compound_pubchem_name = tolower(compound_pubchem_name))

articles <- EIPAAB %>% 
  dplyr::distinct(article_id) %>% 
  nrow(.)

data_points <- EIPAAB %>% 
  nrow(.)

print(paste0("Filtered: n articles = ", articles, "; n data rows = ", data_points))
```
<br>
There are **184 compounds** in this filtered dataset.

```{r}
eipaab_cas_list <- EIPAAB %>% 
  dplyr::distinct(compound_cas) %>% 
  pull(compound_cas)

eipaab_name_list <- EIPAAB %>% 
  dplyr::distinct(compound_name) %>% 
  pull(compound_name)

eipaab_pubchem_name_list <- EIPAAB %>% 
  dplyr::distinct(compound_pubchem_name) %>% 
  pull(compound_pubchem_name)

eipaab_cas_n <- length(eipaab_cas_list)
eipaab_name_n <- length(eipaab_name_list)
eipaab_pubchem_name_n <- length(eipaab_pubchem_name_list)

print(paste0("n CAS numbers = ", eipaab_cas_n, "; n compound names = ", eipaab_name_n, "; n PubChem compound names = ", eipaab_pubchem_name_n))
```

*Note: One of the compounds must have two pubchem names, this isn't an issue, we will use all identifiers to search the environmental occurrence databases*

### EIPAAB tidy

Here we will make a new EIPAAB data frame, **eipaab_tidy**, that we can combind with the other databases more easily. I have added a few new columns so that we can combined it with the environmental occurrence data matrix_group, n_units and source. I have also **filtered** to use only the minimum tested concentration. We have formatted the data so both could be included.

```{r}
eipaab_tidy <- EIPAAB %>% 
  dplyr::select(unique_row_id, compound_cas, compound_name, compound_pubchem_name, compound_min_dose_std, compound_max_dose_std, year, title) %>% 
    pivot_longer(
    cols = c(compound_min_dose_std, compound_max_dose_std),
    names_to = "concentration_type",
    values_to = "value"
  ) %>%
  mutate(concentration_type = case_when(
    concentration_type == "compound_min_dose_std" ~ "min",
    concentration_type == "compound_max_dose_std" ~ "max",
    TRUE ~ NA
  ),
  matrix = "test",
  matrix_group = "test",
  n_units = 1,
  n_units_est = 1,
  source = "eipaab") %>% 
  dplyr::filter(concentration_type == "min") %>% 
  dplyr::rename(reference = title) %>% 
  dplyr::select(source, compound_cas, compound_name, concentration_type, value, matrix_group, matrix, n_units, n_units_est, compound_pubchem_name, year, reference, unique_row_id)
```


## UBA

Import the UBA database, tidy the column names and data. This is the whole UBA database.

```{r}
setwd(input_wd)

column_fix_na <- c("sampling_period_start", "sampling_period_end", "number_of_samples_analysed", "unit_original", "unit_standard", "unit_lo_d", "unit_lo_d_standardized", "lo_d_standardized", "lo_d_limit_of_detection", "sampling_province")

column_fix_chr <- c("matrix", "unit_original", "unit_standard", "unit_lo_d", "unit_lo_d_standardized", "statistics")

UBA <- read_excel("pharms-uba_v3_2021_0.xlsx", skip = 2) %>% 
  janitor::clean_names() %>%
  dplyr::mutate(name_of_analyte = tolower(name_of_analyte),
                across(all_of(column_fix_na), make_na),
                across(all_of(column_fix_chr), tidy_string),
                cas_number = str_replace_all(cas_number, "–", "-"))
```


### Filtering

I am **filtering** for only measures in unit standard of ug/L because we are only comparing water-borne concentrations (i.e. in mass per volume of water). I am also only selecting matrix of interest, and re-allocating the different matrix into broader categories. The matrix of interest that we have selected for this investigation are: <br> 

***Surface waters***: "surface_water_river_stream", "surface_water_unspecific", "surface_water_lake", "surface_water_sea_or_ocean", "surface_water_aquaculture", "surface_water_pond" <br>

***WWTP effluent***: "wwtp_effluent_treated", "wwtp_secondary_effluent", "wwtp_primary_effluent", "wwtp_desinfection_effluent" <br>


```{r}
matrix_surfacewater <- c("surface_water_river_stream", "surface_water_unspecific", "surface_water_lake", "surface_water_sea_or_ocean", "surface_water_aquaculture", "surface_water_pond")

matrix_wwtp_effluent <- c("wwtp_effluent_treated", "wwtp_secondary_effluent", "wwtp_primary_effluent", "wwtp_desinfection_effluent")

matrix_of_intrest <- c(matrix_surfacewater, matrix_wwtp_effluent)


UBA <-  UBA %>% 
  dplyr::filter(unit_standard == "ug_l") %>% 
  dplyr::filter(matrix %in% matrix_of_intrest) %>% 
  dplyr::mutate(matrix_group = case_when(
    matrix %in% matrix_surfacewater ~ "surfacewater",
    matrix %in% matrix_wwtp_effluent ~ "effluent",
    TRUE ~ NA
  ))
```

<br>
The UBA database has measured values as single values or summary statics from a given study. We will highlight this by allocating the different statistics into to measures of central tendency (**central**: "average", "mean", "median", "timeweightedaverage_concentration") vs single values (**single**: "single_value", "single_value_of_a_composite_sample"). There are also summary values such as minimum and maximum, which we will allocate to **"other"**, we will **remove the 'other' data from the database**, and only use the measures of central tendency and single values for concentration comparisons and occurrence. 

```{r}
statistics_central_tendency <- c("average", "mean", "median", "timeweightedaverage_concentration")
statistics_single <- c("single_value", "single_value_of_a_composite_sample")

statistics_select <- c(statistics_single, statistics_central_tendency)

UBA <- UBA %>% 
  dplyr::mutate(statistics_group = case_when(
    statistics %in% statistics_central_tendency ~ "central",
    statistics %in% statistics_single ~ "single",
    TRUE ~ "other"
  )) %>% 
  dplyr::filter(statistics_group != "other")
```

There are **888** distinct CAS numbers in the UBA database and **938** distinct compound names and 122,929 data points for the matrix we are potentially interested in.

```{r}
uba_cas_n <- UBA %>% 
  dplyr::distinct(cas_number) %>% 
  nrow(.)

uba_name_n <- UBA %>% 
  dplyr::distinct(name_of_analyte) %>% 
  nrow(.)

data_points <- UBA %>% nrow()

print(paste0("n CAS numbers = ", uba_cas_n, "; n compound names = ", uba_name_n, "; n data points ", data_points))
```

### UBA tidy

Here we are making a compare data frame to our other databases 

```{r}
uba_tidy <- UBA %>% 
  dplyr::select(id, 
                name_of_analyte, 
                cas_number, 
                matrix, 
                mec_standardized, 
                literature_citation, 
                matrix_group, 
                statistics_group, 
                number_of_samples_analysed,
                sampling_period_end,
                ) %>% 
  dplyr::rename(compound_name = name_of_analyte,
                compound_cas = cas_number,
                unique_row_id = id, 
                concentration_type = statistics_group, 
                value = mec_standardized, 
                n_units = number_of_samples_analysed,
                year = sampling_period_end,
                reference = literature_citation) %>% 
  dplyr::mutate(n_units_est = if_else(is.na(n_units), 1, n_units),
                compound_pubchem_name = NA,
                unique_row_id = as.character(unique_row_id),
                source = "uba") %>% 
  dplyr::select(source, compound_cas, compound_name, concentration_type, value, matrix_group, matrix, n_units, n_units_est, compound_pubchem_name, year, reference, unique_row_id)
```


## Wilkson

Here we import the complete **Wilkson dataset S4** and tidy this dataframe. We have removed all the extra headings, to just have compound name, sample number, and the site headers. We have also filled down compound name.

```{r}
setwd(input_wd)
wilkson <- read_excel("pnas.2113947119.sd04.xlsx", skip = 5) %>% 
  janitor::clean_names() %>% 
  dplyr::slice(6:1103) %>% 
  dplyr::select(-x140) %>% 
  dplyr::rename(compound_name = x1, survey_n = sampling_campaign) %>% 
  dplyr::mutate(compound_name = tolower(compound_name)) %>% 
  tidyr::fill(compound_name)
```

Changing the data to long formate. We will remove empty rows, change the concentration results to numeric, by replacing ND with 0 and removing "e" for estimated values. Lastly, we will change the value to ug/L by dividing by 1000 (it's given in ng/L).

```{r}
wilkson_tidy <- wilkson %>%
  tidyr::pivot_longer(
    cols = -c(compound_name, survey_n), # Columns to keep fixed
    names_to = "site",            # Column to store variable names
    values_to = "value"                # Column to store values
  ) %>%
  dplyr::mutate(
    sample_id = paste(site, survey_n, sep = "_"),
    value = str_remove_all(value, "e"),
    value = case_when(
      is.na(value) ~ NA,
      value == "ND" ~ "0",
      TRUE ~ value
    ),
    value = as.numeric(value)/1000
  ) %>% 
  dplyr::filter(!is.na(value)) %>% 
  dplyr::mutate(unique_row_id = paste0(sample_id, "_", compound_name),
                concentration_type = "single",
                matrix = "river",
                matrix_group = "surfacewater",
                year = 2019, # most samples were collected in 2019 for the project 
                n_units = 1,
                source = "wilkson",
                n_units_est = 1,
                reference = "Wilkson et al 2022",
                compound_pubchem_name = NA,
                compound_cas = NA) %>%
  dplyr::select(-sample_id, -survey_n, -site)  %>% 
  dplyr::select(source, compound_cas, compound_name, concentration_type, value, matrix_group, matrix, n_units, n_units_est, compound_pubchem_name, year, reference, unique_row_id)
```


The Wilkson database has **61 pharmaceuticals** and 64,132 samples

```{r}
compounds <- wilkson_tidy %>% 
  dplyr::distinct(compound_name) %>% 
  nrow(.)

data_points <- wilkson_tidy %>%  nrow(.)

print(paste0("compound names = ", compounds, "; n data points ", data_points))
```


### Filtering

We **do not need to filter** the Wilkson database, all analytes are pharmaceuticals, values are in mass per volume of water, and all samples are from surface waters.


## NORMAN

This part of the code relates to the NORMAN EMPODAT Database - Chemical Occurrence Data (https://www.norman-network.com/nds/empodat/chemicalSearch.php?s=new). This database was searched for all compounds, all countries, all years, all data sources, all organisations, all laboratories, for Surface water (Sea water, River water, Lake water, Transitional water, Coastal water, Territorial (marine) water, Reservoirs, Others) and Waste water (Urban, Industrial, Municipal, Other) Ecosystems/matrices on 19-12-2024 <br>

The complete surface water and effluent NORMAN dataset are very large so I have not uploaded it to GitHub, as a result I have hashed out the below code that was used to import and tidy the complete data set. In a later step I have made a new filtered CSV files called "norman_pham_Surface_water_2024-12-19-214810.csv" and "norman_pham_Waste_water_2025-01-09-222202.csv" which has only compounds of interest. The "norman_pham_Surface_water_2024-12-19-214810.csv" data file is also to large for a standard remote git repository (> 100,000 KB), I have uploaded separately on the OSF framework (DOI: 10.17605/OSF.IO/QDGFJ) please download it from here <br>
<br>
*Download 'norman_pham_Surface_water_2024-12-19-214810.csv': *: https://osf.io/qdgfj/

```{r}
# setwd(input_wd)
# norman_surface_complete <- fread("Surface_water_2024-12-19-214810.csv", skip = 1) %>%
#   clean_names() %>%
#   slice(2:n()) %>%
#   select(sample_matrix, individual_compound, cas_no, norman_sus_dat_id, concentration_data, concentration, value, unit, v34, organisation, references_literature_1) %>%
#   rename(year = v34, compound_name = individual_compound, compound_cas = cas_no, concentration_type = concentration_data, reference = references_literature_1) %>%
#   mutate(compound_name = tolower(compound_name),
#          compound_cas = str_remove_all(compound_cas, "CAS_RN: "),
#          compound_cas = str_replace_all(compound_cas, "–", "-"),
#          matrix_group = "surfacewater",
#          source = "norman",
#          concentration_type = case_when(concentration_type == "Individual results" ~ "single",
#                                         concentration_type == "Aggregate data" ~ "central",
#                                         TRUE ~ concentration_type)
#          )
```

There's 3909 compound CAS numbers (4003 compound names), the earliest record is 1998 and latest is 2020. In total there are 58811806 rows of data (i.e. greater than 58,811,806 samples).

```{r}
# norman_surface_summary <- norman_surface_complete %>% 
#   reframe(n_rows = nrow(.),
#           n_compound_names = length(unique(compound_name)),
#           n_compound_cas = length(unique(compound_cas)),
#           min_year = min(year, na.rm = TRUE),
#           max_year = max(year, na.rm = TRUE)
#             )
# norman_surface_summary
```


Now doing the same for the waste water data

```{r}
# setwd(input_wd)
# norman_waste_complete <- fread("Waste_water_2025-01-09-222202.csv", skip = 1) %>%
#   clean_names() %>%
#   slice(2:n()) %>%
#   select(sample_matrix, individual_compound, cas_no, norman_sus_dat_id, concentration_data, concentration, value, unit, v34, organisation, references_literature_1) %>%
#   rename(year = v34, compound_name = individual_compound, compound_cas = cas_no, concentration_type = concentration_data, reference = references_literature_1) %>%
#   mutate(compound_name = tolower(compound_name),
#          compound_cas = str_remove_all(compound_cas, "CAS_RN: "),
#          compound_cas = str_replace_all(compound_cas, "–", "-"),
#          matrix_group = "waste water",
#          source = "norman",
#          concentration_type = case_when(concentration_type == "Individual results" ~ "single",
#                                         concentration_type == "Aggregate data" ~ "central",
#                                         TRUE ~ concentration_type)
#          )
```

There's 2765 compound CAS numbers (2823 compound names), the earliest record is 2002 and latest is 2019 In total there are 252210 rows of data (i.e. greater than 252,210 samples).

```{r}
# norman_waste_summary <- norman_waste_complete %>%
#   reframe(n_rows = nrow(.),
#           n_compound_names = length(unique(compound_name)),
#           n_compound_cas = length(unique(compound_cas)),
#           min_year = min(year, na.rm = TRUE),
#           max_year = max(year, na.rm = TRUE)
#             )
# norman_waste_summary
```

### Filtering

We will filter this database to only include compound numbers and case from UBA, EIPAAB and Wilkson (i.e. any pharmaceuticals identified across these other three databases). The NORMAN database has all kinds of analytes not just pharmaceuticals, and they are not allocated into compound types/classes, so there is no easy way to independent filter NORMAN for pharmaceutical only. <br>

I have also hashed out parts of this code because we do not want to re-run filtering the the complete NORMAN database.

```{r}
eipaab_name_list_pub <- eipaab_tidy %>%
  dplyr::distinct(compound_pubchem_name) %>%
  rename(compound_name = compound_pubchem_name)

eipaab_name_list <- eipaab_tidy %>%
  dplyr::distinct(compound_name) %>%
  rbind(., eipaab_name_list_pub) %>%
  dplyr::distinct(compound_name)  %>%
  dplyr::pull()

eipaab_cas_list <- eipaab_tidy %>%
  dplyr::distinct(compound_cas)  %>%
  dplyr::pull()

uba_cas_list <- uba_tidy %>%
  dplyr::distinct(compound_cas)  %>%
  dplyr::pull()

uba_name_list <- uba_tidy %>%
  dplyr::distinct(compound_name)  %>%
  dplyr::pull()

wilkson_name_list <- wilkson_tidy %>%
  dplyr::distinct(compound_name) %>%
  dplyr::pull()

all_names_list <- unique(c(eipaab_name_list, uba_name_list, wilkson_name_list))

all_cas_list <- unique(c(eipaab_cas_list, uba_cas_list))
```

<br>
Here we filter the norman surface water data set for these compounds

```{r}
# norman_pham_surface <- norman_surface_complete %>% 
#   dplyr::filter(compound_name %in% all_names_list | compound_cas %in% all_cas_list)
```

<br>
Here we filter the norman wate water data set for these compounds

```{r}
# norman_pham_waste <- norman_waste_complete %>%
#   dplyr::filter(compound_name %in% all_names_list | compound_cas %in% all_cas_list)
```

<br>
Writing these new file to CSV so we can load it from here.

```{r}
setwd(input_wd)
#write.csv(norman_pham_surface, "norman_pham_Surface_water_2024-12-19-214810.csv", row.names = FALSE)
#write.csv(norman_pham_waste, "norman_pham_Waste_water_2025-01-09-222202.csv", row.names = FALSE)
```

<br>
Import the file

```{r}
setwd(input_wd)
norman_pham_surface <- read.csv("norman_pham_Surface_water_2024-12-19-214810.csv")
norman_pham_waste <- read.csv("norman_pham_Waste_water_2025-01-09-222202.csv")
norman_pham <- rbind(norman_pham_surface, norman_pham_waste)
```

<br>
There are 657 compound CAS numbers, 674 compound names and 2,422,683 data points 

```{r}
norman_pham_cas_n <- norman_pham %>%
  dplyr::distinct(compound_cas) %>%
  nrow(.)

norman_pham_name_n <- norman_pham %>%
  dplyr::distinct(compound_name) %>%
  nrow(.)

data_points <- norman_pham %>% nrow()

print(paste0("n CAS numbers = ", norman_pham_cas_n, "; n compound names = ", norman_pham_name_n, "; n data points ", data_points))
```

### NORMAN tidy

```{r}
norman_tidy <- norman_pham %>% 
    dplyr::mutate(n_units = NA, 
                n_units_est = if_else(concentration_type == "central", 2, 1),
                unique_row_id = paste0(norman_sus_dat_id, "_", compound_name),
                compound_pubchem_name = NA,
                matrix = case_when(
                  sample_matrix == "Surface water - River water" ~ "surface_water_river_stream",
                  sample_matrix == "Surface water - Lake water" ~ "surface_water_lake",
                  sample_matrix == "Surface water - Sea water" ~ "surface_water_sea_or_ocean",
                  sample_matrix == "Surface water - Territorial (marine) water" ~ "surface_water_sea_or_ocean",
                  sample_matrix == "Surface water - Transitional water" ~ "surface_water_transitional_water",
                  sample_matrix == "Surface water - Coastal water" ~ "surface_water_sea_or_ocean",
                  sample_matrix == "Surface water - Other" ~ "surface_water_unspecific",
                  sample_matrix == "Surface water - Reservoirs" ~ "surface_water_reservoirs",
                  sample_matrix == "Surface water - Reservoirs" ~ "surface_water_unspecific",
                  sample_matrix == "Waste water - Industrial" ~ "wwtp_industrial_effluent",
                  sample_matrix == "Waste water - Municipal" ~ "wwtp_municipal_effluent",
                  sample_matrix == "Waste water - Other" ~ "wwtp_other_effluent",
                  sample_matrix == "Waste water - Urban" ~ "wwtp_urban_effluent",
                  TRUE ~ "ERROR"
                )) %>%
  dplyr::select(source, compound_cas, compound_name, concentration_type, value, matrix_group, matrix, n_units, n_units_est, compound_pubchem_name, year, reference, unique_row_id)
```

###################
# All databases
###################

## Combind the databases

Here we combined all databases

```{r}
all_databases <- bind_rows(eipaab_tidy, uba_tidy, wilkson_tidy, norman_tidy)
```

## Compound identifies 

In this section we will fix compound synonyms, to make sure names and CAS are consistent across all databases. <br>

First we will identify CAS numbers with multiple names. 

```{r}
cas_with_two_names_list <- all_databases %>% 
  dplyr::filter(!is.na(compound_cas)) %>% 
  dplyr::group_by(compound_cas) %>% 
  dplyr::reframe(names = length(unique(compound_name))) %>% 
  dplyr::filter(names > 1) %>% 
  dplyr::pull(compound_cas)

paste0("There are ", length(cas_with_two_names_list), " CAS with more then one compound name")
```

<br>
We will first reassign them based on the CAS in EIPAAB. Of the 122 CAS with two names, 22 are present in EIPAAB. We will use these to correct the name.

```{r}
eipaab_name_adjust <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  dplyr::filter(compound_cas %in% cas_with_two_names_list) %>% 
  dplyr::group_by(compound_cas) %>% 
  dplyr::slice(1) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(compound_cas, compound_name) %>% 
  dplyr::rename(compound_name_corrected = compound_name)

all_databases <- all_databases %>% 
  dplyr::left_join(., eipaab_name_adjust, by = "compound_cas") %>% 
  dplyr::mutate(compound_name_corrected = if_else(is.na(compound_name_corrected), compound_name, compound_name_corrected))
```

<br>
That fixed 22, but there are still 100 CAS with multiple compound names

```{r}
cas_with_two_names_list <- all_databases %>% 
  dplyr::filter(!is.na(compound_cas)) %>% 
  dplyr::group_by(compound_cas) %>% 
  dplyr::reframe(names = length(unique(compound_name_corrected))) %>% 
  dplyr::filter(names > 1) %>% 
  dplyr::pull(compound_cas)

paste0("There are ", length(cas_with_two_names_list), " CAS with more then one compound name")
```

<br>
Let's select only one of the compound names associated with that CAS number. We will use the most common of the duplicate names.

```{r}
cas_with_two_names <- all_databases %>% 
  dplyr::filter(compound_cas %in% cas_with_two_names_list) %>% 
  dplyr::group_by(compound_cas, compound_name) %>% 
  dplyr::reframe(n = length(compound_cas)) %>% 
  dplyr::arrange(desc(n)) %>% 
  dplyr::group_by(compound_cas) %>% 
  dplyr::slice(1) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(compound_cas, compound_name) %>% 
  dplyr::rename(compound_name_select = compound_name)

all_databases <- all_databases %>% 
  dplyr::left_join(., cas_with_two_names, by = "compound_cas") %>% 
  dplyr::mutate(compound_name_corrected = if_else(is.na(compound_name_select), compound_name_corrected, compound_name_select)) %>% 
  dplyr::select(-compound_name_select)
```

<br>
This fixed the CAS duplicate issue.

```{r}
cas_with_two_names_list <- all_databases %>% 
  dplyr::filter(!is.na(compound_cas)) %>% 
  dplyr::group_by(compound_cas) %>% 
  dplyr::reframe(names = length(unique(compound_name_corrected))) %>% 
  dplyr::filter(names > 1) %>% 
  dplyr::pull(compound_cas)

paste0("There are ", length(cas_with_two_names_list), " CAS with more then one compound name")
```

<br>
Now let's add missing CAS and match CAS to EIPAAB.

<br>
There's 29 compounds in UBA with missing CAS, and 61 in Wilkinson.

```{r}
all_databases %>% 
  dplyr::filter(is.na(compound_cas)) %>% 
  dplyr::group_by(compound_name_corrected) %>% 
  dplyr::slice(1) %>% 
  dplyr::group_by(source) %>% 
  dplyr::reframe(n = length(source))
```

We will use the compound names from EIPAAB to add missing CAS numbers

```{r}
eipaab_idenifiers <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  dplyr::select(compound_cas, compound_name_corrected) %>% 
  dplyr::group_by(compound_cas, compound_name_corrected) %>% 
  dplyr::slice(1) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(compound_cas_corrected = compound_cas)

all_databases <- all_databases %>% 
  dplyr::left_join(., eipaab_idenifiers, by = "compound_name_corrected") %>% 
  dplyr::mutate(compound_cas_corrected = if_else(is.na(compound_cas_corrected), compound_cas, compound_cas_corrected))
```

<br>
Still some with missing CAS

```{r}
all_databases %>% 
  dplyr::filter(is.na(compound_cas_corrected)) %>% 
  dplyr::group_by(compound_name_corrected) %>% 
  dplyr::slice(1) %>% 
  dplyr::group_by(source) %>% 
  dplyr::reframe(n = length(source))
```

<br> 
Adding CAS with Pubchem name

```{r}
eipaab_idenifiers_2 <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  dplyr::select(compound_cas, compound_pubchem_name) %>% 
  dplyr::group_by(compound_cas, compound_pubchem_name) %>% 
  dplyr::slice(1) %>% 
  dplyr::ungroup() %>% 
  dplyr::rename(compound_cas_corrected_2 = compound_cas,
                compound_name_corrected = compound_pubchem_name)

all_databases <- all_databases %>% 
  dplyr::left_join(., eipaab_idenifiers_2, by = "compound_name_corrected") %>% 
  dplyr::mutate(compound_cas_corrected = if_else(is.na(compound_cas_corrected_2), compound_cas_corrected, compound_cas_corrected_2)) %>% 
  dplyr::select(-compound_cas_corrected_2)
```

<br>
There's still 47 compounds in database that don't have a CAS, 28 from UBA and 19 from Wilkson.

```{r}
all_databases %>% 
  dplyr::filter(is.na(compound_cas_corrected)) %>% 
  dplyr::group_by(compound_name_corrected) %>% 
  dplyr::slice(1) %>% 
  dplyr::group_by(source) %>% 
  dplyr::reframe(n = length(source))
```

<br>
Here I have mainly added CAS from PubChem where available for these compound names. 

```{r}
all_databases <- all_databases %>% 
  dplyr::mutate(compound_cas_corrected = case_when(
    compound_name_corrected == "alpha-apo-oxytetracycline" ~ "2869-27-4",
    compound_name_corrected == "artemisinin" ~ "63968-64-9",
    compound_name_corrected == "cloxacillin" ~ "61-72-3",
    compound_name_corrected == "cotinine" ~ "486-56-6",
    compound_name_corrected == "erythrohydrobupropion" ~ "99102-04-2",
    compound_name_corrected == "fluconazole" ~ "86386-73-4",
    compound_name_corrected == "hydrocodone" ~ "125-29-1",
    compound_name_corrected == "itraconazole" ~ "84625-61-6",
    compound_name_corrected == "ketotifen" ~ "34580-13-7",
    compound_name_corrected == "loratadine" ~ "79794-75-5",
    compound_name_corrected == "miconazole" ~ "22916-47-8",
    compound_name_corrected == "nevirapine" ~ "129618-40-2",
    compound_name_corrected == "norethisterone" ~ "68-22-4",
    compound_name_corrected == "norfluoxetine" ~ "83891-03-6",
    compound_name_corrected == "oseltamivir" ~ "196618-13-0",
    compound_name_corrected == "pregabalin" ~ "148553-50-8",
    compound_name_corrected == "raloxifene" ~ "84449-90-1",
    compound_name_corrected == "salbutamol" ~ "18559-94-9",
    compound_name_corrected == "sitagliptin" ~ "486460-32-6",
    compound_name_corrected == "thiabendazole" ~ "148-79-8",
    compound_name_corrected == "triamterene" ~ "396-01-0",
    TRUE ~ compound_name_corrected
  ))
```


## Database numbers

Let's see how much data is present in the filtered and combined databases.

```{r}
total_compounds_eipaab <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  nrow(.)

total_water_samples <- all_databases %>% 
  dplyr::filter(source != "eipaab") %>% 
  dplyr::reframe(n = sum(n_units_est, na.rm = TRUE)) %>% 
  dplyr::pull(n)
  
paste0("There are ", total_compounds_eipaab, " behavioural ecotoxicology exposures and ", total_water_samples, " water samples")
```

```{r}
total_data <- all_databases %>% 
  dplyr::group_by(source) %>% 
  dplyr::reframe(Data = sum(n_units_est, na.rm = TRUE)) %>% 
  dplyr::rename(Source = source) %>% 
  dplyr::mutate(Source = case_when(
    Source == "eipaab" ~ "EIPAAB",
    Source == "norman" ~ "NORMAN*",
    Source == "uba" ~ "PHARMS-UBA",
    Source == "wilkson" ~ "Wilkson et al.",
  ))

total_data %>% 
  gt() %>% 
    fmt_number(
    columns = "Data",
    decimals = 0
    ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )
```
*Note: the NORMAN database was filtered to include only compounds fund in the other databases, so this is not a true representation of the total number of pharmaceutical samples present in the NORMAN database.*

<br>
Number of compounds in each database

```{r}
n_compounds <- all_databases %>% 
  dplyr::group_by(source) %>% 
  dplyr::reframe("Number of compounds" = length(unique(compound_name_corrected))) %>% 
  dplyr::rename(Source = source) %>% 
  dplyr::mutate(Source = case_when(
    Source == "eipaab" ~ "EIPAAB",
    Source == "norman" ~ "NORMAN*",
    Source == "uba" ~ "PHARMS-UBA",
    Source == "wilkson" ~ "Wilkson et al.",
  ))

n_compounds %>% 
    gt() %>% 
    fmt_number(
    columns = "Number of compounds",
    decimals = 0
    ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )
```
*Note: the NORMAN database was filtered to include only compounds fund in the other databases, so this is not a true representation of the total number of different pharmaceuticals present in the NORMAN database.*

#####################
# Analysis 
####################

# Cross-over

```{r}
eipaab_compounds_list <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  dplyr::distinct(compound_name_corrected) %>% 
  dplyr::pull(compound_name_corrected)

n_eipaab_compounds <- length(eipaab_compounds_list)

enviro__cross_over_compounds_list <- all_databases %>% 
  dplyr::filter(source != "eipaab" & compound_name_corrected %in% eipaab_compounds_list) %>% 
  dplyr::distinct(compound_name_corrected) %>% 
  dplyr::pull(compound_name_corrected)

n_cross_over_compounds <- length(enviro__cross_over_compounds_list)

percent_cross_over = round(n_cross_over_compounds/n_eipaab_compounds*100,2)

paste0(n_cross_over_compounds, " of the ", n_eipaab_compounds, " pharmaceuticals tested in the EIPAAB database (with an environmental motivation) were present in the environmental databasess (i.e. ", percent_cross_over, "%)")
```

# Detections

### Summary statistics

Here we make a summary data frame for each compound, **all_surfacewater**, that includes the number of samples,the number of positive detections (based on single samples, not summary statistics), the rank order of detections, the percentage of positive detections (only for compounds with more then 10 sample units), and the rank order of percentage of positive detection. <br>

Here I make a summary data frame, looking at the occurrence and detection rate of compounds in surface waters

```{r}
all_surfacewater <- all_databases %>% 
  dplyr::filter(source != "eipaab", matrix_group == "surfacewater") %>% 
  dplyr::group_by(compound_name_corrected) %>% 
  dplyr::reframe(combind_samples = sum(n_units_est, na.rm = TRUE),
                 combind_rows = length(unique_row_id),
                 combind_singles = sum(concentration_type == "single", na.rm = TRUE),
                 combind_detection = sum(value > 0, na.rm = TRUE),
                 combind_single_detection = sum(value > 0 & concentration_type == "single", na.rm = TRUE),
                 combind_percent_detection = if_else(combind_singles > 10, (combind_single_detection/combind_singles)*100, NA)
                 ) %>% 
  ungroup() %>% 
  dplyr::mutate(
    combind_rank_samples = rank(-combind_samples, ties.method = "max", na.last = "keep"),
    combind_rank_detection = rank(-combind_detection, ties.method = "max", na.last = "keep"),
    combind_rank_detection_percent = rank(-combind_percent_detection, ties.method = "max", na.last = "keep")
  )
```

There's a total of 870 compounds in this surface water summary

```{r}
all_surfacewater %>% 
  dplyr::distinct(compound_name_corrected) %>% 
  nrow(.)
```

Making occurrence count in EIPAAB

```{r}
all_surfacewater_list <- all_surfacewater %>% 
  dplyr::distinct(compound_name_corrected) %>% 
  dplyr::pull(compound_name_corrected)

eipaab_counts <- all_databases %>% 
  dplyr::filter(source == "eipaab" & compound_name_corrected %in% all_surfacewater_list) %>% 
  dplyr::group_by(compound_name_corrected) %>% 
  dplyr::reframe(eipaab_n = sum(n_units_est, na.rm = TRUE),
                 eipaab_n_log = log(eipaab_n)) %>% 
  ungroup() %>% 
  dplyr::mutate(eipaab_rank = rank(-eipaab_n, ties.method = "max", na.last = "keep")) %>% 
  tidyr::complete(., compound_name_corrected)
```

Combining the data together

```{r}
test_surfacewater <- left_join(all_surfacewater, eipaab_counts, by = "compound_name_corrected") %>% 
  dplyr::mutate(rank_samples_dif = eipaab_rank-combind_rank_samples,
                rank_detection_dif = eipaab_rank-combind_rank_detection,
                rank_detection_percent_dif = eipaab_rank-combind_rank_detection_percent
                ) %>% 
  dplyr::arrange(desc(rank_detection_dif), desc(combind_detection)) %>% 
  dplyr::mutate(rank_detection_dif = 1:nrow(.)) %>% 
  dplyr::arrange(desc(rank_detection_percent_dif), desc(combind_percent_detection)) %>% 
  dplyr::mutate(rank_detection_percent_dif = 1:nrow(.))

uba_rank_detection_max <- test_surfacewater %>% 
  dplyr::arrange(desc(combind_rank_detection)) %>% 
  dplyr::slice(1) %>% 
  dplyr::pull(combind_rank_detection)

uba_rank_detection_percent_max <- test_surfacewater %>% 
  dplyr::arrange(desc(combind_rank_detection_percent)) %>% 
  dplyr::slice(1) %>% 
  dplyr::pull(combind_rank_detection_percent)
```

<br>
Let's look at the 10 most common compounds in EIPAAB versus occurrence and detection frequency in UBA surface waters
UBA samples

```{r}
test_surfacewater %>% 
  dplyr::arrange(desc(eipaab_n)) %>% 
  dplyr::slice(1:10) %>% 
  dplyr::mutate(compound_name = sentence_case(compound_name_corrected),
                combind_percent_detection = round(combind_percent_detection,2)) %>% 
  dplyr::select(compound_name, eipaab_n, combind_samples, combind_detection, combind_percent_detection, eipaab_rank, combind_rank_detection, combind_rank_detection_percent) %>% 
  dplyr::rename(Name = compound_name, "EIPAAB tests" = eipaab_n, "Samples" = combind_samples, "All detections" = combind_detection, "Precent single detections" = combind_percent_detection, "EIPAAB rank (1–870)" = eipaab_rank,
                "Rank total detections (1–870)" = combind_rank_detection, "Rank  percent detections (1–713)*" = combind_rank_detection_percent) %>% 
  gt() %>% 
    fmt_number(
    columns = c("Samples", "All detections"), 
    decimals = 0
    ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )
```

<br>
*Note: the rank for total percentage detections does not range from 1 to the number of compounds (i.e. 1–713) because percentage detections were only calculated where the number of single samples are greater than ten* 

### Visualising

#### Total detections 
There's massive skew in occurrence for specific compounds in both the UBA and EIPAAB database, in terms of modelling it, there is likely going to be issues with overdispersion. We will use a negative binomial distribution.

```{r}
visual_1 <- test_surfacewater %>% 
  dplyr::filter(!is.na(eipaab_n)) %>% 
  dplyr::arrange(rank_detection_dif) %>% 
  dplyr::mutate(is_top5 = rank_detection_dif %in% 1:5) %>% 
  ggplot() +
  geom_point(aes(x = combind_detection, y = eipaab_n, size = combind_samples), alpha = 0.4, shape = 21, color = "#01080A", fill = "#E76A24") +
  geom_text_repel(
    data = test_surfacewater %>% filter(!is.na(eipaab_n), rank_detection_dif %in% 1:5),
    aes(x = combind_detection, y = eipaab_n, label = compound_name_corrected), 
    size = 3, box.padding = 0.7, min.segment.length = 0, hjust = 0, vjust = 5
  ) +
  scale_fill_gradientn(colours = colorspace::diverge_hcl(7)) +
  scale_colour_manual(
    values = c("TRUE" = "#01080A", "FALSE" = "#01080A"), 
    guide = "none"
  ) +
  labs(
    title = "Tests vs detection frequency",
    x = "Detections in Environmental Databases",
    y = "Occurrence in EIPAAB",
    size = "Number of samples"
  ) +
  theme_clean() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  )

visual_1
```
<br>
***Note:*** *the size of the points represents the total number of samples in the environmental databases. The five marked points represent those that are under represented in EIPAAB given the number of detections (i.e. rank EIPAAB - rank environmental detections).*


#### Percentage detection

There's massive skew in occurrence for specific compounds in both the UBA and EIPAAB database, in terms of modelling it, there is likely going to be issues with overdispersion. We will use a negative binomial distribution.

```{r}
visual_2 <- test_surfacewater %>% 
  dplyr::filter(!is.na(eipaab_n)) %>% 
  dplyr::arrange(rank_detection_percent_dif) %>% 
  dplyr::mutate(is_top5 = rank_detection_percent_dif %in% 1:5) %>% 
  ggplot() +
  geom_point(aes(x = combind_percent_detection, y = eipaab_n, size = combind_samples), alpha = 0.4, shape = 21, color = "#01080A", fill = "#FBBC42") +
  geom_text_repel(
    data = test_surfacewater %>% filter(!is.na(eipaab_n), rank_detection_percent_dif %in% 1:5),
    aes(x = combind_percent_detection, y = eipaab_n, label = compound_name_corrected), 
    size = 3, box.padding = 0.7, min.segment.length = 0, hjust = 0, vjust = 5
  ) +
  scale_fill_gradientn(colours = colorspace::diverge_hcl(7)) +
  scale_colour_manual(
    values = c("TRUE" = "#01080A", "FALSE" = "#01080A"), 
    guide = "none"
  ) +
  labs(
    title = "Tests vs percentage of detections",
    x = "Percent Detections in Environmental Databases",
    y = "Occurrence in EIPAAB",
    size = "Number of samples"
  ) +
  theme_clean() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  )

visual_2
```
<br>
***Note:*** *the size of the points represents the total number of samples in the environmental databases. The five marked points represent those that are under represented in EIPAAB given the percentage of positive detections (i.e. rank EIPAAB - rank environmental detections).*


#### Temporal trends

Here we will look at temporal trends using only EIPAAB, UBA and NORMAN data. 

```{r}
temporal_trends <- all_databases %>% 
  dplyr::filter(source != "wilkson") %>% 
  dplyr::mutate(group_type = if_else(source == "eipaab", "test", "environmental")) %>% 
  dplyr::group_by(group_type, year) %>% 
  dplyr::reframe(n_samples = sum(n_units_est, na.rm = TRUE),
                 n_compounds = length(unique(compound_name_corrected))) %>% 
  tidyr::complete(group_type, year) %>% 
  dplyr::filter(!is.na(year)) %>% 
  dplyr::group_by(group_type) %>% 
  dplyr::mutate(n_samples_z = scale(n_samples)) %>% 
  dplyr::ungroup()
```
 
Let's look at the total number of samples per year

```{r}
temporal_trends %>% 
  dplyr::filter(year > 1999 & year < 2021) %>% 
  ggplot(aes(x = year, y = n_samples)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~group_type, scales = "free",
             labeller = labeller(group_type = c(
               "environmental" = "Environmental data",
               "test" = "Behavioural ecotoxicology data")) 
             ) +
    labs(
    title = "Data over time",
    x = "Year",
    y = "Data points (tests or water samples)"
    ) +
  theme_clean() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  )
```

Now the number of different compounds per year

```{r}
temporal_trends %>% 
  dplyr::filter(year > 1999 & year < 2021) %>% 
  ggplot(aes(x = year, y = n_compounds, size = n_samples_z)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~group_type, scales = "free",
             labeller = labeller(group_type = c(
               "environmental" = "Environmental data",
               "test" = "Behavioural ecotoxicology data")) 
             ) +
    labs(
    title = "Number of compounds over time",
    x = "Year",
    y = "Data points (tests or water samples)"
    ) +
  theme_clean() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  )
```


```{r}
eipaab_compounds_list <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  dplyr::distinct(compound_name_corrected) %>% 
  dplyr::pull(compound_name_corrected)

temporal_trends_by_compound <- all_databases %>% 
  dplyr::filter(source != "wilkson" & compound_name_corrected %in% eipaab_compounds_list) %>% 
  dplyr::mutate(group_type = if_else(source == "eipaab", "test", "environmental")) %>% 
  dplyr::group_by(group_type, year, compound_name_corrected) %>% 
  dplyr::reframe(n_samples = sum(n_units_est, na.rm = TRUE)) %>% 
  tidyr::complete(group_type, year, compound_name_corrected) %>% 
  dplyr::filter(!is.na(year))

temporal_trends_enviro <- temporal_trends_by_compound %>% 
  dplyr::filter(group_type == "environmental") %>% 
  dplyr::select(-group_type, -year, -compound_name_corrected) %>% 
  dplyr::rename(n_samples_env = n_samples)

temporal_trends_test <- temporal_trends_by_compound %>% 
  dplyr::filter(group_type == "test") %>% 
  dplyr::select(-group_type) %>% 
  dplyr::rename(n_samples_test = n_samples)

temporal_trends_cbind <- cbind(temporal_trends_test, temporal_trends_enviro)
```


Looking at compound by compound sample occurrence over time across the environmental and EIPAAB databases. There's no obvious relationship. Perhaps there would be a time lag we could look for? 

```{r}
temporal_trends_cbind %>% 
  dplyr::filter(year > 2008 & year < 2020) %>% 
  ggplot(aes(x = n_samples_test, y = n_samples_env, color = compound_name_corrected)) +
  geom_point(alpha = 0.5) +
  facet_wrap(~year, scales = "free") +
  guides(color = "none") 
```


### Modeling

#### Total detections

Here we will build a simple Bayesian negative binomial (log-link) regression model to look at the relationship between the two parameters. <br>

```{r}
detection_negbin_str <- bf(eipaab_n ~ combind_detection,
               family = negbinomial())
```

These are the default priors <br>

```{r}
suppressWarnings(get_prior(detection_negbin_str, data = test_surfacewater, family = negbinomial()))
```

Run model. This has been hashed out as I have saved the model, and will re-load it instead of re-running it. <br>

```{r}
setwd(output_mod_wd)

# detection_negbin_mod <- brm(detection_negbin_str,
#                data = test_surfacewater,
#                cores = 4,
#                chains = 4,
#                #prior = mod_priors, #use defaults
#                warmup = 1000,
#                seed = 20250418,
#                thin = 2,
#                iter = 8000,
#                control = list(max_treedepth = 20, adapt_delta = 0.9999),
#                save_pars = save_pars(all=TRUE),
#                sample_prior = TRUE,
#                file = 'detection_negbin_mod')
print("Model complete")
```

##### Re-load models

Reload the models if necessary

```{r}
setwd(output_mod_wd)

detection_negbin_mod <-  readRDS(file = "detection_negbin_mod.rds")
```


##### Model diagnostics

The R2 value is pretty low, but should be fine for the purposes of this investigation.

```{r}
performance::r2_bayes(detection_negbin_mod, robust = FALSE, ci = 0.95)
```

The mode generally follows the observed data, but isn't the best fit. Again, this should be fine for the purposes of this investigation

```{r}
brms::pp_check(detection_negbin_mod, ndraws = 50, type = "ecdf_overlay")
```

Checking R-hat values to confirm model convergence. 


```{r}
print(summary(detection_negbin_mod, prob = 0.95)) #All Rhat = 1 (good).
```

Diagnostic plots look fine. A smaller shape value indicates greater overdispersion (variance is much larger than the mean).

```{r}
plot(detection_negbin_mod)
```
##### Model estimates

Pulling the conditional effect estimates from the model. The estimate corresponds to the effect of the predictor on the log of the expected response variable. A 1-unit increase (i.e. 1 detection) in the number of environmental detections increases the expected occurrence in EIPAAB database by 0.005% (i.e. (exp(4.950996e-05)-1)*100 = 0.004951119). In other words, for each 1000 detections we would expect a 5% increase in test in the EIPAAB database.

```{r}
# Extract estimates
esitmates_1 <- conditional_effects(detection_negbin_mod)
estimate_data_1 <- esitmates_1$combind_detection

# Extract the summary as a data frame
model_summary_1 <- as.data.frame(summary(detection_negbin_mod, prob = 0.95)$fixed)

estimate_1 <- model_summary_1[2,1]
L95CI_1 <- model_summary_1[2,3]
H95CI_1 <- model_summary_1[2,4]
estimate_percent_1 <- (exp(estimate_1)-1)*100
L95CI_percent_1 <- (exp(L95CI_1)-1)*100
H95CI_percent_1 <- (exp(H95CI_1)-1)*100
```


#### Percentage detections

```{r}
percent_detection_negbin_str <- bf(eipaab_n ~ combind_percent_detection,
               family = negbinomial())
```

Run model. This has been hashed out as I have saved the model, and will re-load it instead of re-running it. <br>

```{r}
setwd(output_mod_wd)

# percent_detection_negbin_mod <- brm(percent_detection_negbin_str,
#                data = test_surfacewater,
#                cores = 4,
#                chains = 4,
#                #prior = mod_priors, #use defaults
#                warmup = 1000,
#                seed = 20250418,
#                thin = 2,
#                iter = 8000,
#                control = list(max_treedepth = 20, adapt_delta = 0.9999),
#                save_pars = save_pars(all=TRUE),
#                sample_prior = TRUE,
#                file = 'percent_detection_negbin_mod')
print("Model complete")
```

##### Re-load models

Reload the models if necessary

```{r}
setwd(output_mod_wd)

percent_detection_negbin_mod <-  readRDS(file = "percent_detection_negbin_mod.rds")
```


##### Model diagnostics

The R2 value is pretty low, but should be fine for the purposes of this investigation

```{r}
performance::r2_bayes(percent_detection_negbin_mod, robust = FALSE, ci = 0.95)
```

```{r}
brms::pp_check(percent_detection_negbin_mod, ndraws = 100, type = "ecdf_overlay")
```

Checking R-hat values to confirm model convergence. 

```{r}
print(summary(percent_detection_negbin_mod, prob = 0.95)) #All Rhat = 1 (good).
```

```{r}
plot(percent_detection_negbin_mod)
```

##### Model estimates

Pulling the conditional effect estimates from the model. The estimate corresponds to the effect of the predictor on the log of the expected response variable. A 1-unit increase (i.e. 1%) in the percentage of environmental detections increases the expected occurrence in EIPAAB database by 1.17% (i.e. (exp(0.01160477)-1)*100 = 1.167237).

```{r}
# Extract estimates
esitmates_2 <- conditional_effects(percent_detection_negbin_mod)
estimate_data_2 <- esitmates_2$combind_percent_detection

# Extract the summary as a data frame
model_summary_2 <- as.data.frame(summary(percent_detection_negbin_mod, prob = 0.95)$fixed)

estimate_2 <- model_summary_2[2,1]
L95CI_2 <- model_summary_2[2,3]
H95CI_2 <- model_summary_2[2,4]
estimate_percent_2 <- (exp(estimate_2)-1)*100
L95CI_percent_2 <- (exp(L95CI_2)-1)*100
H95CI_percent_2 <- (exp(H95CI_2)-1)*100
```

### Results 

#### Total detections

There is generally a small positive correlation between the total environmental detections in the surface water data and occurrence in the EIAPAAB database (Fig 1; Bayesian negative binomial regression estimate: 4.950996e-05 [1.490186e-05, 9.115236e-05]). With that said, there are a number of compounds that occurrence frequently in the surface waters that are not well represented in the EIPAAB database (see Table 3 below). These could be sensible targets for future behavioural tests. <br>

##### Fig 1a

**Fig 1a:** the relationship between the number of surface water detections for pharmaceuticals in environmental databases (Norman, UBA and Wilkson) and the occurrence in the Evidence of the Impacts of Pharmaceuticals on Aquatic Animal Behaviour (EIPAAB) database.

```{r}
fig_1a <- ggplot() +
  geom_line(data = estimate_data_1, aes(x = combind_detection, y = estimate__), color = "#1c4595", size = 1) +
  geom_ribbon(data = estimate_data_1, aes(x = combind_detection, ymin = lower__, ymax = upper__), fill = "#1c4595", alpha = 0.1) +
  geom_point(data = test_surfacewater %>% filter(!is.na(eipaab_n)) %>% dplyr::arrange(rank_detection_dif) %>% 
               mutate(is_top5 = rank_detection_dif %in% 1:5), aes(x = combind_detection, y = eipaab_n, size = combind_samples), alpha = 0.4, shape = 21, color = "#01080A", fill = "#E76A24") +
  # Add text annotation
  annotate(
    "text",
    x = max(test_surfacewater$combind_detection, na.rm = TRUE)/2, 
    y = max(test_surfacewater$eipaab_n, na.rm = TRUE), 
    label = paste0(
      "Bayesian negative binomial regression", "\n",
      "Effect estimate (%): ", round(estimate_percent_1, 3), "\n",
      "CI: ", round(L95CI_percent_1, 3), " - ", round(H95CI_percent_1, 3)
    ),
    hjust = 0, vjust = 1, size = 4
  ) +
  geom_text_repel(
    data = test_surfacewater %>% filter(!is.na(eipaab_n), rank_detection_dif %in% 1:5),
    aes(x = combind_detection, y = eipaab_n, label = compound_name_corrected), size = 3, box.padding = 0.7
  ) +
  # Adjust scales
  scale_fill_gradientn(colours = colorspace::diverge_hcl(7)) +
  scale_colour_manual(
    values = c("TRUE" = "#01080A", "FALSE" = "#01080A"), guide = "none"
    ) +
  labs(
    title = "Test vs detection frequency",
    x = "Detections in Environmental Databases",
    y = "Occurrence in EIPAAB",
    fill = "Rank Difference",
    size = "Rank Difference"
  ) +
  theme_clean() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  )

fig_1a
```

<br>
***Note:*** *the size of the points represents the total number of samples in the environmental databases. The five marked points represent those that are under represented in EIPAAB given the number of detections (i.e. rank EIPAAB - rank environmental detections).*

<br>
This code saves the figure above as PDF

```{r}
ggsave(filename = paste0(output_fig_wd, "./figure-1a.pdf"), 
       plot = fig_1a,
       width = 210,
       height = 297/1.5,  # Specify the height of the plot
       units = "mm")
```

##### Table 1
<br>
**Table 1:** Compounds with the most environmental detections that are not present in EIPAAB.

```{r}
table_1 <- test_surfacewater %>% 
  dplyr::filter(is.na(eipaab_n)) %>% 
  dplyr::arrange(desc(combind_samples)) %>% 
  dplyr::slice(1:15) %>% 
  dplyr::mutate(compound_name = sentence_case(compound_name_corrected)) %>% 
  dplyr::select(compound_name, combind_samples, combind_detection, combind_percent_detection, combind_rank_samples, combind_rank_detection) %>% 
  dplyr::rename(Name = compound_name, "Samples" = combind_samples, "Detections" = combind_detection, "Percentage detections" = combind_percent_detection, "Rank samples" = combind_rank_samples, "Rank detections" = combind_rank_detection) %>% 
  gt() %>% 
  fmt_number(
    columns = c("Samples", "Detections"), 
    decimals = 0
  ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )

table_1
```

```{r}
gtsave(table_1, filename = paste0(output_fig_wd, "./table-1.png"))
```


#### Percentage detections

##### Fig 1b
**Fig 1b:** the relationship between the percentage of surface water detections for pharmaceuticals in environmental databases (Norman, UBA and Wilkson) and the occurrence in the Evidence of the Impacts of Pharmaceuticals on Aquatic Animal Behaviour (EIPAAB) database.


```{r}
fig_1b <- ggplot() +
  geom_line(data = estimate_data_2, aes(x = combind_percent_detection, y = estimate__), color = "#1c4595", size = 1) +
  geom_ribbon(data = estimate_data_2, aes(x = combind_percent_detection, ymin = lower__, ymax = upper__), fill = "#1c4595", alpha = 0.1) +
  geom_point(data = test_surfacewater %>% filter(!is.na(eipaab_n)) %>% dplyr::arrange(rank_detection_percent_dif) %>% 
               mutate(is_top5 = rank_detection_percent_dif %in% 1:5), 
             aes(x = combind_percent_detection, y = eipaab_n, size = combind_singles), alpha = 0.4, shape = 21, color = "#01080A", fill = "#fbbc42") +
  annotate(
    "text",
    x = 40, 
    y = max(test_surfacewater$eipaab_n, na.rm = TRUE), 
    label = paste0(
      "Bayesian negative binomial regression", "\n",
      "Effect estimate (%): ", round(estimate_percent_2, 3), "\n",
      "CI: ", round(L95CI_percent_2, 3), " - ", round(H95CI_percent_2, 3)
    ),
    hjust = 0, vjust = 1, size = 4
  ) +
  geom_text_repel(
    data = test_surfacewater %>% filter(!is.na(eipaab_n), rank_detection_percent_dif %in% 1:5),
    aes(x = combind_percent_detection, y = eipaab_n, label = compound_name_corrected), size = 3, box.padding = 0.7
    ) +
  scale_fill_gradientn(colours = colorspace::diverge_hcl(7)) +
  scale_colour_manual(
    values = c("TRUE" = "#01080A", "FALSE" = "#01080A"), guide = "none") +
  labs(
    title = "Test vs Percentage Detections",
    x = "Percentage detections in Environmental Databases",
    y = "Occurrence in EIPAAB",
    size = "Number of single samples"
  ) +
  theme_clean() +
  theme(
    legend.position = "bottom",
    legend.justification = "center"
  )

fig_1b
```

##### Table 2

**Table 2:** The compounds with the highest detection rate not found in the EIPAAB database

```{r}
table_2 <- test_surfacewater %>%
  dplyr::filter(is.na(eipaab_n) & !is.na(combind_percent_detection)) %>%
  dplyr::arrange(desc(combind_percent_detection)) %>% 
  dplyr::slice(1:15) %>% 
  dplyr::mutate(compound_name = sentence_case(compound_name_corrected)) %>% 
  dplyr::select(compound_name, combind_singles, combind_single_detection, combind_percent_detection) %>% 
  dplyr::rename(Name = compound_name, "Single samples" = combind_singles, "Single detections" = combind_single_detection, "Percent detections" = combind_percent_detection) %>% 
  gt() %>% 
  fmt_number(
    columns = c("Percent detections"), 
    decimals = 2
  ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )

table_2
```

```{r}
gtsave(table_2, filename = paste0(output_fig_wd, "./table-2.png"))
```

## Concentration 

For questions around concentration, I will filtered for only cases where the levels are above zero or above the limit of detection (i.e. positive detections or above LOD). We will do this for two reasons, the first being that when researchers are trying to estimate the risk of environmental exposure, we assume that they are emulating potential exposure events at a contaminated site, not trying to replicate a theoretical global surface water average. The second is, by removing zero values and values below LODs, we are trying to be ***more conservative*** with the relative comparison between tested doses in behavioural ecotoxicology and observed environmental doses (as opposed to using some value substitute/estimated value for samples below detection limit). We are also using both single values and summary values of central tendency (means, median, ect). <br>

In this section we are calculating summary metrics for concentrations used in the behavioural test and detected in surface waters and effluent in the enviromental data. We will do this separately for effluent and surface waters samples. We are calculating, the mean, median, min, max, range, n (number of contributing data points), standard deviation (sd), standard error (se), and the upper and lower 95% credible intervals (ci_upper, ci_lower).

```{r}
conc_summary <- all_databases %>% 
  dplyr::filter(value > 0) %>% 
  dplyr::group_by(matrix_group, compound_name_corrected) %>% 
  dplyr::reframe(mean = mean(value, na.rm = TRUE),
                 median = median(value, na.rm = TRUE),
                 min = min(value, na.rm = TRUE),
                 max = max(value, na.rm = TRUE),
                 range = max-min,
                 n = length(unique_row_id),  # Sample size
                 sd = sd(value, na.rm = TRUE),  # Standard deviation
                 ci_lower = quantile(value, 0.025, na.rm = TRUE),  #95% credible interval using quantiles
                 ci_upper = quantile(value, 0.975, na.rm = TRUE)
                 ) %>% 
  dplyr::mutate(ci_lower = if_else(is.na(ci_lower), NA, ci_lower),
                ci_upper = if_else(is.na(ci_upper), NA, ci_upper)) %>% 
  tidyr::complete(
    tidyr::expand(nesting(matrix_group, compound_name_corrected))
  )
```

<br>
Here we will add the summary statistics from above to the full database, so we can see how many EIPAAB tested concentrations fall under the median surface water and effluent concentrations for each compound.

```{r}
rename_cols <- conc_summary %>% 
  dplyr::select(-matrix_group, -compound_name_corrected) %>% 
  colnames()

conc_summary_eff <- conc_summary %>% 
  dplyr::filter(matrix_group == "effluent") %>% 
   dplyr::rename_with(
    .cols = all_of(rename_cols),  # Select columns to rename
    .fn = ~ paste0("effluent_", .)  # Prefix with "effluent_"
  ) %>% 
  dplyr::select(-matrix_group)


conc_summary_surf <- conc_summary %>% 
  dplyr::filter(matrix_group == "surfacewater") %>% 
   dplyr::rename_with(
    .cols = all_of(rename_cols),  # Select columns to rename
    .fn = ~ paste0("surfacewater_", .)  # Prefix with "effluent_"
  ) %>% 
  dplyr::select(-matrix_group)


all_databases_with_sum <- all_databases %>% 
  dplyr::filter(source == "eipaab") %>% 
  dplyr::left_join(., conc_summary_eff, by = "compound_name_corrected") %>% 
  dplyr::left_join(., conc_summary_surf, by = "compound_name_corrected") %>% 
  dplyr::mutate(under_med_surf = if_else(value < surfacewater_median, 1, 0),
                under_hci_surf = if_else(value < surfacewater_ci_upper, 1, 0),
                under_med_effl = if_else(value < effluent_median, 1, 0),
                under_hci_effl = if_else(value < effluent_ci_upper, 1, 0),
                diff_med_surf = value/surfacewater_median,
                diff_hic_surf = value/surfacewater_ci_upper,
                diff_med_effl = value/effluent_median,
                diff_hic_effl = value/effluent_ci_upper
                )
```

<br>
Here we are summarising how many test were below the median surface water and effluent concentrations as well as the upper 95%CI. We are also calculating the mean-fold difference in concentrations.

```{r}
conc_summary <- all_databases_with_sum %>% 
  dplyr::reframe(eipaab_n = length(unique_row_id),
                 n_sw = sum(!is.na(under_med_surf)),
                 less_med_sw = sum(under_med_surf, na.rm = TRUE),
                 less_hci_sw = sum(under_hci_surf, na.rm = TRUE),
                 percent_less_med_sw = (less_med_sw/n_sw)*100,
                 percent_less_hic_sw = (less_hci_sw/n_sw)*100,
                 n_ef = sum(!is.na(under_med_effl)),
                 less_med_ef = sum(under_med_effl, na.rm = TRUE),
                 less_hci_ef = sum(under_hci_effl, na.rm = TRUE),
                 percent_less_med_ef = (less_med_ef/n_ef)*100,
                 percent_less_hic_ef = (less_hci_ef/n_ef)*100,
                 fdiff_med_sw = median(diff_med_surf, na.rm = TRUE),
                 fdiff_hci_sw = median(diff_hic_surf, na.rm = TRUE),
                 fdiff_med_ef = median(diff_med_effl, na.rm = TRUE),
                 fdiff_hci_ef = median(diff_hic_effl, na.rm = TRUE),
                 ) %>% 
  dplyr::mutate(across(everything(), ~ ifelse(is.nan(.), NA, .)))
```

#### Results

For the X compounds in the EIPAAB database that had associated data in UBA (representing 1,440 exposures), only X% and X% employed test concentrations lower then median and upper 95% credible interval of surface water concentrations, respectively (Table 4). Further, only X% and X% employed test concentrations lower then median and upper 95% credible interval of wastewater treatment plant effluent (Table 4). On average, concentrations used in behavioural ecotoxicology were X times higher then median surf water concentrations, and X times higher then those in effluent. <br>

##### Table 3

**Table 3:** The number of exposures in the EIPAAB database that used concentrations less then median and upper 95% credible interval of surface water and effluent concentrations

```{r}
surf <- conc_summary %>% 
  dplyr::select(n_sw, less_med_sw, less_hci_sw, fdiff_med_sw, fdiff_hci_sw, percent_less_med_sw, percent_less_hic_sw) %>% 
  dplyr::rename(total_n = n_sw, less_med = less_med_sw, less_hci = less_hci_sw, fdiff_med = fdiff_med_sw, fdiff_hci = fdiff_hci_sw, percent_less_med = percent_less_med_sw, percent_less_hic = percent_less_hic_sw) %>% 
  dplyr::mutate(matrix = "Surface water")

effluent <- conc_summary %>% 
  dplyr::select(n_ef, less_med_ef, less_hci_ef, fdiff_med_ef, fdiff_hci_ef, percent_less_med_ef, percent_less_hic_ef) %>% 
  dplyr::rename(total_n = n_ef, less_med = less_med_ef, less_hci = less_hci_ef, fdiff_med = fdiff_med_ef, fdiff_hci = fdiff_hci_ef, percent_less_med = percent_less_med_ef, percent_less_hic = percent_less_hic_ef) %>% 
  dplyr::mutate(matrix = "Effluent")

table_3 <- rbind(surf, effluent) %>% 
  dplyr::select(matrix, everything()) %>% 
  dplyr::mutate(percent_less_med = round(percent_less_med, 2),
                percent_less_hic = round(percent_less_hic, 2),
                fdiff_med = round(fdiff_med, 2),
                fdiff_hci = round(fdiff_hci, 2)) %>% 
  dplyr::rename(Matrix = matrix, 'Total tests' = total_n, 'Dose less than median' = less_med, 'Dose less than upper 95%CI' = less_hci, "Mean-fold differene from median" = fdiff_med, "Mean-fold differene from upper 95%CI" = fdiff_hci, 'Percent less than median (%)' = percent_less_med, 'Percent less than upper 95%CI (%)' = percent_less_hic) %>% 
  gt() %>% 
  fmt_number(
    columns = 'Total tests', 
    decimals = 0
  ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )

table_3
```
Saving the table 

```{r}
gtsave(table_3, filename = paste0(output_fig_wd, "./table-3.png"))
```

<br>
Now we will make a summary for each compound

```{r}
sample_counts <- all_databases_with_sum %>% 
  dplyr::select(compound_name_corrected, surfacewater_n, effluent_n) %>% 
  group_by(compound_name_corrected) %>% 
  dplyr::slice(1) %>% 
  ungroup()

under_levels <- all_databases_with_sum %>% 
  dplyr::group_by(compound_name_corrected) %>% 
  dplyr::reframe(eipaab_n = length(unique_row_id),
                 n_sw = sum(!is.na(under_med_surf)),
                 less_med_sw = sum(under_med_surf, na.rm = TRUE),
                 less_hci_sw = sum(under_hci_surf, na.rm = TRUE),
                 prop_less_med_sw = less_med_sw/n_sw,
                 prop_less_hic_sw = less_hci_sw/n_sw,
                 n_ef = sum(!is.na(under_med_effl)),
                 less_med_ef = sum(under_med_effl, na.rm = TRUE),
                 less_hci_ef = sum(under_hci_effl, na.rm = TRUE),
                 prop_less_med_ef = less_med_ef/n_ef,
                 prop_less_hic_ef = less_hci_ef/n_ef,
                 fdiff_med_sw = median(diff_med_surf, na.rm = TRUE),
                 fdiff_hci_sw = median(diff_hic_surf, na.rm = TRUE),
                 fdiff_med_ef = median(diff_med_effl, na.rm = TRUE),
                 fdiff_hci_ef = median(diff_hic_effl, na.rm = TRUE),
                 ) %>% 
  dplyr::left_join(., sample_counts, by = "compound_name_corrected") %>% 
  dplyr::mutate(across(everything(), ~ ifelse(is.nan(.), NA, .)))
```

##### Fig 2a

**Fig 2a.** The median fold difference in tested concentrations in EIPAAB and median surface water concentration (i.e. tested dose / median surface water concentration)

```{r, fig.height=15, fig.width=5}
fdiff_med_sw_order <- under_levels %>% 
  dplyr::filter(!is.na(fdiff_med_sw)) %>% 
  dplyr::arrange(desc(fdiff_med_sw)) %>% 
  dplyr::mutate(order = 1:nrow(.),
                compound_name_corrected = sentence_case(compound_name_corrected)) %>% 
  pull(compound_name_corrected)

vline_data <- data.frame(xintercept = log(10^(0:9)),
                         xlabs = as.character(10^(0:9))) %>% 
  dplyr::mutate(xlabs = paste0("×",xlabs))

median <- conc_summary %>% 
  dplyr::pull(fdiff_med_sw) %>% 
  log()

fig_2a <- under_levels %>% 
  dplyr::filter(!is.na(fdiff_med_sw)) %>% 
  dplyr::mutate(compound_name_corrected = sentence_case(compound_name_corrected),
                compound_name_corrected = factor(compound_name_corrected, levels = rev(fdiff_med_sw_order))) %>% 
  ggplot(aes(x = log(fdiff_med_sw), y = compound_name_corrected, size = eipaab_n, fill = log(surfacewater_n))) +
  geom_vline(data = vline_data, aes(xintercept = xintercept), linetype = "dashed", colour = "black") +
  geom_vline(xintercept = median, color = "darkred") +
  geom_text(data = vline_data, aes(x = xintercept, y = "Risperidone", label = xlabs), inherit.aes = FALSE, angle = 90, vjust = -0.5, hjust = 1, size = 3) +
  geom_point(alpha = 0.5, shape = 21) +
  theme_classic() +
  scale_fill_gradientn(colours = colorspace::diverge_hcl(7)) +
  theme(
    legend.position = "bottom",  # Move legend to the top
    legend.justification = "center",  # Align legend to the left
  ) +
  labs(
    subtitle = "Tested concentrations vs median surfwater water",
    x = "Log mean fold difference between tested and median surfwater water concentrations",
    y = "Compounds",
    size = "EIPAAB data",
    fill = "Environmental data"
  ) +
    theme(
    legend.position = "bottom",  # Move legend to the top
    legend.justification = "center"  # Align legend to the left
  )

fig_2a
```

<br>
***Note:*** *the size of the points represent the number of occurrence in EIPAAB, and colour of the point represents the number of detections in the environmental databases* 

<br>
Save the figure

```{r}
ggsave(filename = paste0(output_fig_wd, "./figure-2a.pdf"), 
       plot = fig_2a,
       width = 210/2,
       height = 297,  # Specify the height of the plot
       units = "mm")
```

##### Fig 2b

**Fig 2a.** The median fold difference in tested concentrations in EIPAAB and median effluent concentration (i.e. tested dose / median effluent concentration)

```{r, fig.height=15, fig.width=5}
fdiff_med_ef_order <- under_levels %>% 
  dplyr::filter(!is.na(fdiff_med_ef)) %>% 
  dplyr::arrange(desc(fdiff_med_ef)) %>% 
  dplyr::mutate(order = 1:nrow(.),
                compound_name_corrected = sentence_case(compound_name_corrected)) %>% 
  pull(compound_name_corrected)

vline_data <- data.frame(xintercept = log(10^(0:9)),
                         xlabs = as.character(10^(0:9))) %>% 
  dplyr::mutate(xlabs = paste0("×",xlabs))

median <- conc_summary %>% 
  dplyr::pull(fdiff_med_ef) %>% 
  log()

fig_2b <- under_levels %>% 
  dplyr::filter(!is.na(fdiff_med_ef)) %>% 
  dplyr::mutate(compound_name_corrected = sentence_case(compound_name_corrected),
                compound_name_corrected = factor(compound_name_corrected, levels = rev(fdiff_med_ef_order))) %>% 
  ggplot(aes(x = log(fdiff_med_ef), y = compound_name_corrected, size = eipaab_n, fill = log(surfacewater_n))) +
  geom_vline(data = vline_data, aes(xintercept = xintercept), linetype = "dashed", colour = "black") +
  geom_vline(xintercept = median, color = "darkred") +
  geom_text(data = vline_data, aes(x = xintercept, y = "Lorazepam", label = xlabs), inherit.aes = FALSE, angle = 90, vjust = -0.5, hjust = 1, size = 3) +
  geom_point(alpha = 0.5, shape = 21) +
  theme_classic() +
  scale_fill_gradientn(colours = colorspace::diverge_hcl(7)) +
  theme(
    legend.position = "bottom",  # Move legend to the top
    legend.justification = "center",  # Align legend to the left
  ) +
  labs(
    subtitle = "Tested concentrations vs median surfwater water",
    x = "Log mean fold difference between tested and median surfwater water concentrations",
    y = "Compounds",
    size = "EIPAAB data",
    fill = "Environmental data"
  ) +
    theme(
    legend.position = "bottom",  # Move legend to the top
    legend.justification = "center"  # Align legend to the left
  )

fig_2b
```

<br>
***Note:*** *the size of the points represent the number of occurrence in EIPAAB, and colour of the point represents the number of detections in the environmental databases* 

<br>
Save the figure

```{r}
ggsave(filename = paste0(output_fig_wd, "./figure-2b.pdf"), 
       plot = fig_2b,
       width = 210/2,
       height = 297,  # Specify the height of the plot
       units = "mm")
```

<br>

Looking at the number of compounds in the database that have data at concentrations below median and upper 95 CI

##### Table 5

**Table 5:** The percentage of compounds in the EIPAAB database that have data at concentrations less then median and upper 95% credible interval of surface water and effluent concentrations

```{r}
compound_sum_table_surf <- under_levels %>% 
  dplyr::reframe(total_compounds =  sum(!is.na(prop_less_med_sw)),
                 count_less_med_sw = sum(prop_less_med_sw == 0, na.rm = TRUE),
                 count_less_hic_sw = sum(prop_less_hic_sw == 0, na.rm = TRUE),
                 percent_less_med_sw = (count_less_med_sw/total_compounds)*100,
                 percent_less_hic_sw = (count_less_hic_sw/total_compounds)*100,
                 "Percent with no data under median concentration" = paste0(round(percent_less_med_sw, 2), "%", " (", count_less_med_sw, " of ", total_compounds, ")"),
                 "Percent with no data under upper 95 CI concentration" = paste0(round(percent_less_hic_sw, 2), "%", " (", count_less_hic_sw, " of ", total_compounds, ")")
                 ) %>% 
  dplyr::mutate(Matrix = "Surface water") %>% 
  dplyr::select(Matrix,  "Percent with no data under median concentration", "Percent with no data under upper 95 CI concentration")

compound_sum_table_eff <- under_levels %>% 
  dplyr::reframe(total_compounds =  sum(!is.na(prop_less_med_ef)),
                 count_less_med_ef = sum(prop_less_med_ef == 0, na.rm = TRUE),
                 count_less_hic_ef = sum(prop_less_hic_ef == 0, na.rm = TRUE),
                 percent_less_med_ef = (count_less_med_ef/total_compounds)*100,
                 percent_less_hic_ef = (count_less_hic_ef/total_compounds)*100,
                 "Percent with no data under median concentration" = paste0(round(percent_less_med_ef, 2), "%", " (", count_less_med_ef, " of ", total_compounds, ")"),
                 "Percent with no data under upper 95 CI concentration" = paste0(round(percent_less_hic_ef, 2), "%", " (", count_less_hic_ef, " of ", total_compounds, ")")
                 ) %>% 
  dplyr::mutate(Matrix = "Effluent") %>% 
  dplyr::select(Matrix,  "Percent with no data under median concentration", "Percent with no data under upper 95 CI concentration")

compound_sum_table <-  rbind(compound_sum_table_surf, compound_sum_table_eff)

compound_sum_table %>% 
  gt() %>% 
  fmt_number(
    columns = c("Percent with no data under median concentration", "Percent with no data under upper 95 CI concentration"), 
    decimals = 0
  ) %>% 
  cols_align(
    align = "center", 
    columns = everything()
  )
```


